---
output:
  html_document: default
  pdf_document: default
---
# Exploring White Wine Quality by Ursula Neumann
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(GGally)
library(memisc)
library(gridExtra)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the data
df <- read.csv('white_wine_quality.csv', check.names = FALSE, row.names = 1)
head(df)
```

# Introduction
This dataset, originally described by 
[Cortez et al. 2009](https://www.sciencedirect.com/science/article/pii/S0167923609001377) 
contains 4,898 white wines with 11 variables quantifying the chemical 
properties of each wine. At least 3 wine experts rated the quality of each 
wine, providing a rating between 0 (very bad) and 10 (very excellent).
[This link](https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt)
describes of the variables and how the data was collected.  I am interested in
investigating which chemical properties influence the quality of white wines.

# Univariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Dimensions}
# Determine the dimensions of the dataset
dim(df)
```

As expected there are 4898 white wines, with 12 columns (11 physicochemical
properties and 1 quality rating).

```{r echo=FALSE, message=FALSE, warning=FALSE, Structure}
# Determine the structure of the dataset
str(df)
```

All of the 11 physicochemical properties are of the numeric type (contains 
decimals) while the quality values are integers. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Summary}
# Provide a summary of the dataset
summary(df)
```

For all variables the median and the mean are close together suggesting that 
these parameters will not be highly skewed.  The median (5.2) and mean (6.3) of 
the residual sugar variable are the farthest apart suggesting that there may be 
some outliers on the high end (for example the max of 65.8).  It is surprising 
to observe a fold difference of over 100 between the minimum and maximum values 
for residual sugar and free sulfur dioxide.  The alcohol content ranges from 
8.0 - 14.2%. The scale for quality was 1-10; however, the ratings given only 
range from 3-9, with a median rating of 6.

```{r echo=FALSE, message=FALSE, warning=FALSE, Histogram_Fixed_Acidity}
# Plot a histogram of fixed acidity levels
ggplot(aes(x = fixed.acidity),
       data = df) +
  geom_histogram()
```

The histogram of fixed acidity appears to be normally distributed.

```{r echo=FALSE, message=FALSE, warning=FALSE, Histogram_Volative_Acidity}
# Plot a histogram of volatile acidity levels
ggplot(aes(x = volatile.acidity),
       data = df) +
  geom_histogram()
```

The histogram of volatile acidity is slightly skewed to the right.
I will log10 transform this data to see if it makes the histogram
normally distributed.

```{r echo=FALSE, message=FALSE, warning=FALSE, Histogram_Volative_Acidity_Log}
# Plot a histogram of volatile acidity levels on log10 scale
ggplot(aes(x = volatile.acidity),
       data = df) +
  geom_histogram() +
  scale_x_log10()
```

Now volatile acidity is normally distributed.

```{r echo=FALSE, message=FALSE, warning=FALSE, Histogram_Citric_Acid}
# Plot a histogram of citric acid levels
ggplot(aes(x = citric.acid),
       data = df) +
  geom_histogram()
```

The bulk of the data of citric acid is normally distributed with an outlier of 
1.66 pulling the skew to the right.

```{r echo=FALSE, message=FALSE, warning=FALSE, Histogram_Citric_Acid_0.99}
# Plot a histogram of the citric acid levels excluding the top 1%
ggplot(aes(x = citric.acid),
       data = df) +
  geom_histogram() +
  xlim(0, quantile(df$citric.acid, probs = 0.99))
```

When the top 1% of citric acid data is removed the data appear normally 
distributed. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Histogram_Residual_Sugar}
# Plot a histogram of residual sugar levels
ggplot(aes(x = residual.sugar),
       data = df) +
  geom_histogram()
```

The histogram of residual sugar is skewed to the right.  This graph has the bin
with the highest count near the left with a wide right tail and an outlier at 
65.8.

```{r echo=FALSE, message=FALSE, warning=FALSE, Histogram_Residual_Sugar_Log}
# Plot a histogram of residual sugar levels on a log10 scale
ggplot(aes(x = residual.sugar),
       data = df) +
  geom_histogram() +
  scale_x_log10()
```

Transforming the data using a log10 scale does not make the data normally
distributed, but bimodal.

```{r echo=FALSE, message=FALSE, warning=FALSE, Histogram_Chlorides}
# Plot a histogram of chloride levels
ggplot(aes(x = chlorides),
       data = df) +
  geom_histogram()
```

The histogram of chlorides is skewed to the right with a long thin tail.

```{r echo=FALSE, message=FALSE, warning=FALSE, Histogram_Chlorides_Log}
# Plot a histogram of chloride levels on a log10 scale
ggplot(aes(x = chlorides),
       data = df) +
  geom_histogram() +
  scale_x_log10()
```

Transforming the chlorides data using log10 makes the data appear
more normally distributed.

```{r echo=FALSE, message=FALSE, warning=FALSE, Histogram_Free_Sulfur_Dioxide}
# Plot a histogram of free sulfur dioxide levels
ggplot(aes(x = free.sulfur.dioxide),
       data = df) +
  geom_histogram()
```

The histogram of free sulfur dioxide levels appears to be slightly skewed
to the right with a very small and thin tail and an outlier at 289.

```{r echo=FALSE, message=FALSE, warning=FALSE, Histogram_Free_Sulfur_Dioxide_2}
# Plot a histogram of free sulfur dioxide levels excluding the top 1%
ggplot(aes(x = free.sulfur.dioxide),
       data = df) +
  geom_histogram() +
  xlim(0, quantile(df$free.sulfur.dioxide, probs = 0.99))
```

When excluding the top 1% of free sulfur dioxide levels, the data appear more
symmetric, however, the data looks more like a pyramid instead of normally 
distributed.

```{r echo=FALSE, message=FALSE, warning=FALSE, Histogram_Total_Sulfur_Dioxide}
# Plot a histogram of total sulfur dioxide levels
ggplot(aes(x = total.sulfur.dioxide),
       data = df) +
  geom_histogram()
```

The histogram of total sulfur dioxide generally looks normally distributed
with an extremely thin right tail.

```{r echo=FALSE, message=FALSE, warning=FALSE, Histogram_Density}
# Plot a histogram of density levels
ggplot(aes(x = density),
       data = df) +
  geom_histogram()
```

The histogram of density appears to be normally distributed with a very thin 
right tail and an outlier at 1.039.  The majority of densities are close to, but
below, 1 which makes sense given the density of water being 1 and alcohol being 
less than 1.  I'm unsure what would cause a wine to have a density above 1.

```{r echo=FALSE, message=FALSE, warning=FALSE, Histogram_pH}
# Plot a histogram of pH levels
ggplot(aes(x = pH),
       data = df) +
  geom_histogram()
```

The histogram of pH appears to be normally distributed.

```{r echo=FALSE, message=FALSE, warning=FALSE, Histogram_Sulphates}
# Plot a histogram of sulphate levels
ggplot(aes(x = sulphates),
       data = df) +
  geom_histogram()
```

The histogram of sulphates appears to be skewed to the right.

```{r echo=FALSE, message=FALSE, warning=FALSE, Histogram_Sulphates_Log}
# Plot a histogram of sulphate levels on a log10 scale
ggplot(aes(x = sulphates),
       data = df) +
  geom_histogram() +
  scale_x_log10()
```

Graphing sulphates on a log10 scale transforms the data to be normally 
distributed.

```{r echo=FALSE, message=FALSE, warning=FALSE, Histogram_Alcohol}
# Plot a histogram of alcohol levels
ggplot(aes(x = alcohol),
       data = df) +
  geom_histogram()
```

The histogram of alcohol levels is skewed to the right with a wide tail.

```{r echo=FALSE, message=FALSE, warning=FALSE, Histogram_Alcohol_Log}
# Plot a histogram of alcohol levels on a log10 scale
ggplot(aes(x = alcohol),
       data = df) +
  geom_histogram() +
  scale_x_log10()
```

Graphing the alcohol data on a log10 scale makes the data more symmetric but 
does not appear normally distributed as the peak of the data is wide.

```{r echo=FALSE, message=FALSE, warning=FALSE, Histogram_Quality}
# Plot a histogram of quality levels
ggplot(aes(x = quality),
       data = df) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = seq(3, 9, 1))
```

The quality rating with the highest count is 6 and the data are largely 
symmetric as the ratings for 5 and 7 are slightly lower, 4 and 8 are even lower,
3 and 9 are very low, and 0, 1, 2 and 10 are absent.

# Univariate Analysis

### What is the structure of your dataset?
The dataset contains data on 4898 white wines with 12 columns 
(11 physicochemical properties and 1 quality rating). All of the 
11 physicochemical properties are of the numeric type (contains 
decimals) while the quality values are integers. 

### What is/are the main feature(s) of interest in your dataset?
I would like to investigate which physicochemical properties of the wine 
influence the quality rating; therefore, the quality rating is the most 
important feature in this data.  

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
I will investigate all 11 properties of the wines to determine whether they
influence quality.  Given my limited knowledge of wine and these specific
physicochemical properties, I cannot predict which features will affect
quality.  In addition, the univariate analyses alone does not shed light
on which variables will be the most important therefore I will investigate
them all in the bivariate section.

### Did you create any new variables from existing variables in the dataset?
I have not created any new variables yet.  This may change after further 
investigation into the bi- and multi-variate analyses.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
The data was already tidy therefore no adjustments were made.


# Bivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Scatterplot_Matrix}
# Create a scatterplot matrix to investigate relationships between variables
ggpairs(data=df, size=0.5)
```

Although I am most interested in determining if any of these physicochemical
properties influence quality, we will investigate all relationships first to
better understand the data.  I will report on correlations that are equal to 
or greater than |0.3| when rounded to 1 decimal point with correlations between 
|0.3| and |0.7| to be considered moderate and above |0.7| to be strong.

fixed acidity vs citric acid 0.3
fixed acidity vs pH -0.4

free sulfur dioxide vs residual sugar 0.3
free sulfur dioxide vs total sulfur dioxide 0.6

density vs fixed acidity 0.3
density vs residual sugar 0.8
density vs chlorides 0.3
density vs free sulfur dioxide 0.3
density vs total sulfur dioxide 0.5

alcohol vs density -0.8
alcohol vs residual sugar -0.5
alcohol vs chlorides -0.4
alcohol vs free sulfur dioxide -0.3
alcohol vs total sulfur dioxide -0.5

quality vs alcohol 0.4
quality vs density -0.3

It is logical that fixed acidity would be positively correlated to citric acid 
and negatively correlated to pH given each that each parameter is related to 
acidity.

As expected free and total sulfur dioxide levels are strongly positively 
correlated to one another since total levels are equal to free levels + bound 
levels of sulfur dioxide.

Interestingly density and alcohol were correlated to many other parameters.
It makes sense that density would increase with increasing solutes (residual 
sugar, chlorides, etc.) but I'm assuming that because density decreases with 
increasing alcohol levels (given the density of alcohol vs water) this explains 
why alcohol is therefore negatively correlated to these solutes as well.

The only parameters that are related to quality are alcohol (which are
moderately positive correlated), and density (which are moderately negative
correlated).

```{r echo=FALSE, message=FALSE, warning=FALSE, Scatterplot_Alcohol_Quality}
# Plot a scatterplot of quality vs alcohol
ggplot(aes(x = alcohol,
           y = quality),
       data = df) +
  geom_point(alpha = 1/10) +
  stat_smooth(method = 'lm')
```

Graphing the data with quality on a continuous scale I am able to add a linear 
fit line through the data showing that as alcohol levels increase, quality 
levels increase. However, despite the transparency of the points, it is 
difficult to visualize the extent of the overlap between each quality group.
Therefore, I have graphed same data using jittering to see more points, as well 
as see the 5 point summary in a boxplot as shown below.

```{r echo=FALSE, message=FALSE, warning=FALSE, Boxplot_Alcohol_Quality}
# Add a column for quality as a decrete variable to use in a boxplot
df$quality.descrete <- cut(df$quality, seq(0,10,1), labels=c(1:10))
head(df)

# Plot a boxplot of alcohol and quality
p1 <- ggplot(aes(x = quality.descrete,
           y = alcohol),
       data = df) +
  geom_jitter(aes(color = quality.descrete))

p2 <- ggplot(aes(x = quality.descrete,
           y = alcohol),
       data = df) +
  geom_boxplot(aes(color = quality.descrete))


grid.arrange(p1, p2, nrow = 1)
```

In the graph on the left we can see that the majority of the data falls in
quality rating 5, 6 and 7, less data falls in a rating of 4 and 8, even less so 
in the rating of 3, and only five points in a rating of 9.  There is a great 
deal of overlap between each sequential rating.

Interestingly, in the graph on the right, it looks like the alcohol rating
tends to decrease from a quality rating of 3 to 5, and increases from 5 to 9.
The group with the least variability is the quality rating of 9 where 80% of
the points are clustered together at a high alcohol rating.  However, there is  
a very low N (only 5 observations in this category), and variability may 
increase with more observations.

```{r echo=FALSE, message=FALSE, warning=FALSE, Scatterplot_Density_Quality}
# Plot a scatterplot of quality vs density levels
ggplot(aes(x = density,
           y = quality),
       data = df) +
  geom_point() 
```

It looks like the 2 outliers with a density above 1.01 are making it more
difficult to view the bulk of the data.  I will exclude these points and graph 
the plot again.

```{r echo=FALSE, message=FALSE, warning=FALSE, Scatterplot_Density_Quality2}
# Plot a scatterplot of quality vs density levels excluding 2 outliers
ggplot(aes(x = density,
           y = quality),
       data = subset(df, df$density < 1.01)) +
  geom_point(alpha = 0.1) +
  stat_smooth(method = 'lm') 
```

Applying the linear model line on the data we can see that with increasing 
density there is a decrease in quality.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Boxplot_Density_Quality}
# Plot a boxplot of density and quality
p3 <- ggplot(aes(x = quality.descrete,
           y = density),
       data = subset(df, df$density < 1.01)) +
  geom_jitter(aes(color = quality.descrete))

p4 <- ggplot(aes(x = quality.descrete,
           y = density),
       data = subset(df, df$density < 1.01)) +
  geom_boxplot(aes(color = quality.descrete))


grid.arrange(p3, p4, nrow = 1)
```

Similar to the alcohol vs quality plots, the density vs quality plots also 
exhibit significant overlap between the quality groups. Because density and 
alcohol are related (increased alcohol leads to decreased density), again we 
observe the least variability in the group with a quality rating of 9 where 80% 
of the points are clustered together at a low density level.  

So far it does not appear that either alcohol or density (or anything else) are 
strong indicators of quality.  Therefore, to investigate the strongest 
correlation observed I will graph density vs alcohol. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Scatterplot_Alcohol_Density}
# Plot a scatterplot of quality vs density levels excluding 2 outliers
ggplot(aes(x = alcohol,
           y = density),
       data = subset(df, df$density < 1.01)) +
  geom_point() +
  stat_smooth(method = 'lm') 
```

This indeed is a stronger relationship.  Again this is not surprising given
that alcohol is less dense than water.  However, I anticipate that predicting
density (rather than qulity) will end up being more fruitful using predictive
models, albeit potentially less interesting.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Using the scatterplot matrix I was able to quickly observe the correlations
between all variables.  It was not surprising that fixed acidity was positively 
correlated to citric acid and negatively correlated to pH given each that each 
parameter is related to acidity.  In addition, free and total sulfur dioxide 
levels are strongly positively correlated to one another since total levels are 
equal to free levels + bound levels of sulfur dioxide.

Unfortunately, alcohol and density were only weakly to moderately correlated to 
quality (0.4 and -0.3 respectively), and therefore I do not anticipate that the 
predictive model I will generate in the next section will be very accurate.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
What was more interesting to me was the relationship between density and 
alcohol, as well as the relationship between each of these parameters to fixed 
acidity, residual sugar, chlorides, and total sulfur dioxide levels. I 
anticipate that the predictive model generated for predicting density will be 
more accurate than the one predicting quality.

### What was the strongest relationship you found?
The strongest relationships found were density vs residual sugar (0.8) and 
density vs alcohol (-0.8).


# Multivariate Plots Section

I am first going to attempt to make a predictive model for quality.  Quality was 
most strongly correlated to alcohol and I would like to see if any other 
variables help explain the variability seen in the quality vs alcohol graph.

```{r echo=FALSE, message=FALSE, warning=FALSE, Density_Bucket}
# Create a new column bucketing the density levels into Low, Medium, and High
df$density.bucket <- cut(df$density, 
                         breaks = c(-Inf, 0.9925, 0.995, Inf),
                         labels = c("Low", "Medium", "High"))
summary(df$density.bucket)
```

For all instances of me dividing up variables into 3 roughly equal-sized 
groups, I chose the values manually as I was curious to play around with the 
numbers; however, I believe there is also a easy to cut them into 3 equal groups 
automatically.

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_Alcohol_Density}
# Plot a scatterplot of quality vs alcohol colored by density levels
ggplot(aes(x = alcohol,
           y = quality),
       data = subset(df, df$density < 1.01)) +
  geom_point(aes(color = density.bucket)) +
  scale_color_manual(values=c("#9ecae1", "#4292c6", "#084594")) +
  stat_smooth(method = 'lm')
```

Density appears to be more correlated to alcohol levels alone, opposed to 
explaining the variability in this graph. Next, I will look at the residual 
sugar level, which was highly correlated to density (which was moderately 
correlated to quality).

```{r echo=FALSE, message=FALSE, warning=FALSE, Residual_Sugar__Bucket}
# Create a new column bucketing the residual sugar levels
df$residual.sugar.bucket <- cut(df$residual.sugar, 
                         breaks = c(-Inf, 2, 8, Inf),
                         labels = c("Low", "Medium", "High"))
summary(df$residual.sugar.bucket)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_Alcohol_Residual_Sugar}
# Plot a scatterplot of quality vs alcohol colored by residual sugar levels
ggplot(aes(x = alcohol,
           y = quality),
       data = subset(df, df$density < 1.01)) +
  geom_point(aes(color = residual.sugar.bucket)) +
  scale_color_manual(values=c("#9ecae1", "#4292c6", "#084594")) +
  geom_smooth(method = 'lm')
```

The different colored points (varied residual sugar levels) appear to be equally
distributed throughout the plot and therefore likely do not explain the 
variability observed.

For curiosity's sake I will generate a predictive model incorporating alcohol 
first, then adding on other variables to see their effect on the R-squared value
of the model.  However, I anticipate that the model will be very poor.

```{r echo=FALSE, message=FALSE, warning=FALSE, Predictive_Model_Quality}
# Create a linear predictive model for quality
m1 <- lm(I(quality) ~ I(alcohol), data = df)
m2 <- update(m1, ~ . + density)
m3 <- update(m1, ~ . + residual.sugar)
m4 <- update(m1, ~ . + chlorides)
m5 <- update(m2, ~ . + residual.sugar)
m6 <- update(m5, ~ . + chlorides)
mtable(m1, m2, m3, m4, m5, m6)
```

m1 using alcohol alone as a predicter elicited a R-squared value of only 0.190.
Adding on density, residual sugar, and chlorides alone increased the R-squared 
value minutely.  m6 using all of these parameters had the highest R-squared 
value of 0.212 which is still a very poor model.  Again, out of curiosity I 
wanted to determine how large the 95% confidence intervals would be when 
predicting a wine with the median value for all variables, as well as a wine 
with a quality rating of 9.

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_Estimite_1}
# Predict quality of a wine with median levels of each property
MedianWine <- data.frame(alcohol = 10.4, 
                      density = 0.9937,
                      residual.sugar = 5.2, 
                      chlorides = 0.043)
modelEstimate1 <- predict(m6, newdata = MedianWine,
                        interval="prediction", level = .95)
modelEstimate1
```

For the median wine which would have a rating of ~6, the estimate of 5.8 is not 
that far off.

```{r echo=FALSE, message=FALSE, warning=FALSE, Highest_Rated_Wines_1}
# Display the properties of the highest rated wines
head(subset(df, df$quality == 9))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_Estimite_2}
# Predict the quality of wine #775 which was rated to have a quality of 9
HighestRatedWine <- data.frame(alcohol = 10.4, 
                      density = 0.9970,
                      residual.sugar = 10.6, 
                      chlorides = 0.035)
modelEstimate2 <- predict(m6, newdata = HighestRatedWine,
                        interval="prediction", level = .95)
modelEstimate2
```

When qualities of a highly rated (quality 9) wine are inputted into the model, 
the estimate is identical to before, a value of only 5.8 and a value of 9 did
not fall within the 95% confidence interval.  I expect for any parameters that 
are inputted into the model, it will predict the same quality and thus is a very
poor model.

Moving away from quality, would like to try and predict the density of wine 
based on the alcohol levels, as well as additional parameters starting with 
residual sugar.

```{r echo=FALSE, message=FALSE, warning=FALSE, Density_Alcohol_Residual_Sugar}
# Plot a scatterplot of density vs alcohol colored by residual sugar levels
ggplot(aes(x = alcohol,
           y = density),
       data = subset(df, df$density < 1.01)) +
  geom_point(aes(color = residual.sugar.bucket)) +
  stat_smooth(method = 'lm') +
  scale_color_manual(values=c("#9ecae1", "#4292c6", "#084594"))
```

Residual sugar appears to explain some of the variability in the density vs 
alcohol graph since the points above the linear line of best fit have the 
highest sugar levels, and the observations below the line contain the lowest 
sugar levels. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Chlorides_Bucket}
# Create a new column bucketing the chlorides levels
df$chlorides.bucket <- cut(df$chlorides, 
                         breaks = c(0, 0.038, 0.048, Inf),
                         labels = c("Low", "Medium", "High"))
summary(df$chlorides.bucket)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Density_Alcohol_Chlorides}
# Plot a scatterplot of density vs alcohol colored by chloride levels
ggplot(aes(x = alcohol,
           y = density),
       data = subset(df, df$density < 1.01)) +
  geom_point(aes(color = chlorides.bucket)) +
  stat_smooth(method = 'lm') +
  scale_color_manual(values=c("#9ecae1", "#4292c6", "#084594"))
```

Chlorides do not seem to explain the variability in the density vs alcohol plot 
since the color shift seems to be more related to alcohol levels alone.

```{r echo=FALSE, message=FALSE, warning=FALSE, Free_Sulfur_Dioxide_Bucket}
# Create a new column bucketing the free sulfur dioxide levels
df$free.sulfur.dioxide.bucket <- cut(df$free.sulfur.dioxide, 
                         breaks = c(0, 25, 40, Inf),
                         labels = c("Low", "Medium", "High"))
summary(df$free.sulfur.dioxide.bucket)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Density_Alcohol_Free_Sul_Diox}
# Plot a scatterplot of density vs alcohol colored by free sulfur dioxide levels
ggplot(aes(x = alcohol,
           y = density),
       data = subset(df, df$density < 1.01)) +
  geom_point(aes(color = free.sulfur.dioxide.bucket)) +
  stat_smooth(method = 'lm') +
  scale_color_manual(values=c("#9ecae1", "#4292c6", "#084594"))
```

I do not observe any pattern when bucketing by free sulfur dioxide levels.

```{r echo=FALSE, message=FALSE, warning=FALSE, Total_Sulfur_Dioxide_Bucket}
# Create a new column bucketing the total sulfur dioxide levels
df$total.sulfur.dioxide.bucket <- cut(df$total.sulfur.dioxide, 
                         breaks = c(0, 115, 155, Inf),
                         labels = c("Low", "Medium", "High"))
summary(df$total.sulfur.dioxide.bucket)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Density_Alcohol_Total_Sul_Diox}
# Plot a scatterplot of density vs alcohol colored by total sulfur dioxide
ggplot(aes(x = alcohol,
           y = density),
       data = subset(df, df$density < 1.01)) +
  geom_point(aes(color = total.sulfur.dioxide.bucket)) +
  stat_smooth(method = 'lm') +
  scale_color_manual(values=c("#9ecae1", "#4292c6", "#084594"))
```

It is difficult to say whether total sulfur dioxide levels explain the 
variability in the density vs alcohol plot. There is a cluster of "high" points 
on the top left above the linear line of best fit, however, the "low" points 
seem to congregate at higher alcohol levels both below and above the linear 
line.

```{r echo=FALSE, message=FALSE, warning=FALSE, Fixed_Acidity_Bucket}
# Create a new column bucketing the fixed acidity levels
df$fixed.acidity.bucket <- cut(df$fixed.acidity, 
                         breaks = c(0, 6.4, 7, Inf),
                         labels = c("Low", "Medium", "High"))
summary(df$fixed.acidity.bucket)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Density_Alcohol_Fixed_Acidity}
# Plot a scatterplot of density vs alcohol colored by fixed acidity levels
ggplot(aes(x = alcohol,
           y = density),
       data = subset(df, df$density < 1.01)) +
  geom_point(aes(color = fixed.acidity.bucket)) +
  stat_smooth(method = 'lm') +
  scale_color_manual(values=c("#9ecae1", "#4292c6", "#084594"))
```

Fixed acidity levels may explain some of the variability in the plot of density 
vs alcohol since the highest points at any given alcohol level tend to be darker
(higher fixed acidity) while the lowest points tend to be the lightest (lowest 
fixed acidity).

Next I will generate a predictive model of density.

```{r echo=FALSE, message=FALSE, warning=FALSE, Predictive_Model_Density}
# Create a linear predictive model for density
m1 <- lm(I(density) ~ I(alcohol), data = df)
m2 <- update(m1, ~ . + residual.sugar)
m3 <- update(m2, ~ . + fixed.acidity)
m4 <- update(m3, ~ . + chlorides)
m5 <- update(m4, ~ . + free.sulfur.dioxide)
m6 <- update(m5, ~ . + total.sulfur.dioxide)
mtable(m1, m2, m3, m4, m5, m6)

# After running the models above it seems like alcohol, residual sugar, fixed 
# acidity, and total sulfur dioxide levels will best predict density levels 
# (since chlorides and free sulfur dioxide levels did not increase the R-squared
# value)
m7 <- update(m3, ~ . + total.sulfur.dioxide)
mtable(m1, m2, m3, m4, m5, m6, m7)
```

Alcohol as a single variable produced an R-squared value of 0.609 when
explaining the density levels.  Adding residual sugar, fixed acidity, and total 
sulfur dioxide levels into the equation increased the R-squared value to 0.907, 
0.930, and 0.933 respectively.  There may be parameters that were not measured 
the explain the remainder of the variability in density levels.  This R-squared 
value is much higher than the predictive model for quality, and I anticipate 
that it will be able to accurately predict density levels which is tested below.

```{r echo=FALSE, message=FALSE, warning=FALSE, Highest_Rated_Wines_2}
# Display the properties of the highest rated wines
head(subset(df, df$quality == 9))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Density_Estimate}
# Predict the density of wine #821 (actual density is 0.997)
Wine821 <- data.frame(alcohol = 12.4, 
                      residual.sugar = 1.6,
                      fixed.acidity = 6.6, 
                      total.sulfur.dioxide = 85)
modelEstimate3 <- predict(m7, newdata = Wine821,
                        interval="prediction", level = .95)
modelEstimate3
```

Surprisingly despite the R-squared value of 0.933, the true density of the
wine (0.997) didn't land within the 95% confidence interval.  Perhaps I put too 
much hope in this model because the R-squared value was so much higher than that
of the model predicting quality, but the R-squared value may need to be closer 
to 1 to better predict density.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?
For the multivariate analysis of quality, I did not find anything useful beyond 
alcohol and density that could explain quality levels, and this resulted in a 
very poor predictive model.  Interestingly, I found that a combination of 
alcohol and residual sugar levels can largely explain density levels (and fixed 
acidity and total sulfur dioxide levels help the prediction slightly more.)

### Were there any interesting or surprising interactions between features?
See above.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
The predictive model of quality was poor.  None of the individual parameters
were strongly correlated to quality, and no combination that I tested added to 
the quality of the prediction. 

The predictive model for density was much better than quality with an R-squared 
value of 0.933, however this was still unable to predict the density of a wine 
within the 95% confidence interval.  There must be other parameters that explain
density that were not included in this dataset. 

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
# Plot a scatterplot of quality vs alcohol
ggplot(aes(x = alcohol,
           y = quality),
       data = df) +
  geom_point(alpha = 1/10) +
  stat_smooth(method = 'lm') +
  ggtitle('Quality Increases as Alcohol Levels Increase') +
  theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Alcohol Level (% by volume)") +
  ylab("Quality Rating")
```

### Description One
Firstly, I wanted variety in the appearance of my final plots so I chose this 
one even though it may be unusual to graph a scatterplot when the quality 
ratings are descrete.  I thought this graph was important because it represents 
the strongest correlation between any variable and quality which ended up being 
alcohol.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
ggplot(aes(x = quality.descrete,
           y = density),
       data = subset(df, df$density < 1.01)) +
  geom_boxplot(aes(color = quality.descrete)) +
  ggtitle('Density Decreases as Quality Increases') +
  theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Quality Rating") +
  ylab(expression(paste("Density (g/cm"^3, ")"))) + 
  theme(legend.position="none")
```

### Description Two
I chose this graph as it was a different way to represent the type of data shown
in Plot 1.  The boxplot allows you to glean information such as the 5 number 
summary, outliers, overlap between groups, and trends across quality ratings.  

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
# Plot a scatterplot of density vs alcohol colored by residual sugar levels
ggplot(aes(x = alcohol,
           y = density),
       data = subset(df, df$density < 1.01)) +
  geom_point(aes(color = residual.sugar.bucket)) +
  stat_smooth(method = 'lm') +
  scale_color_manual(values=c("#9ecae1", "#4292c6", "#084594")) +
  ggtitle('Residual Sugar Explains Variability of Density vs Alcohol Plot') +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x="Alcohol Level (% by volume)", 
       y=expression(paste("Density (g/cm"^3, ")")), 
       col="Residual Sugar Level")
```

### Description Three
The strongest relationships found were between density vs alcohol (-0.8) and 
density vs residual sugar (0.8) and this multivariate plot was able to display 
all three variables at once.  It appears that residual sugar levels explain some
of the variability in the density vs alcohol plot, and alcohol and residual 
sugar alone were able to predict density levels with an R-squared value of 
0.907.

------

# Reflection

I found that plotting all the histograms in the univariate plots was tedious, 
boring, and not very informative since it is not intuitive what all the 
variables mean, or the significance of the values.  

I was disappointed that I wasn't able to find anything that strongly correlated
with the quality of wine, and that the predictive model was so poor. For 
something like the diamonds dataset, it is more intuitive that you'd be able
to predict the price of a diamond - generally, the bigger, the better.  However, 
with wine, quality can be quite subjective, especially for an expert vs a 
non-expert. Some people love sweet and some people hate it.  Some people may be 
turned off if the alcohol taste is too strong. Some experts might be swayed by 
hints of flavors and depth of taste while I just hope the wine doesn't taste 
like cleaner.  

After doing a more in-depth investigation into what dictates density I was a 
little more excited and willing to complete this project.  Although it is 
obvious that alcohol levels and solutes will affect density, it was fun to play 
with the predictions using the linear models.

For future analysis, I think it would be interesting if use more complicated 
modelling techniques to create more accurate predictive models such as those 
used in this [R Pubs post](http://www.rpubs.com/datascientiest/237448).

# Resources

https://stackoverflow.com/questions/6286313/remove-an-entire-column-from-a-data-frame-in-r
https://www.sciencedirect.com/science/article/pii/S0167923609001377
https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt 
https://s3.amazonaws.com/content.udacity-data.com/courses/ud651/diamondsExample_2016-05.html
http://colorbrewer2.org/#type=sequential&scheme=Blues&n=7
https://stackoverflow.com/questions/40675778/center-plot-title-in-ggplot2 
https://stackoverflow.com/questions/35618260/removing-legend-ggplot-2-2
http://www.rpubs.com/datascientiest/237448